Comparing Machine Learning algorithms for classifying cognitive states from fMRI data

### 2012-11-09 article/article.tex
abstract

%% Abstract section.
\abstract{
Functional magnetic resonance imaging (fMRI) is a brain mapping technique that allows researchers to measure activity in a subject's brain. This technique is typically used to determine which areas of the brain are involved in performing a certain task. Using this knowledge, we may also be able to determine what task was performed by looking at fMRI data. In this paper, we will discuss the use of machine learning algorithms for the classification of fMRI data. By using classifiers we will try to predict the cognitive state of subjects, based on their brain activity during that time.
} % end of abstract

@@

In this study we use the same StarPlus data as in the previously mentioned experiment, and use machine learning algorithms to decode the cognitive states of subjects when they are looking at pictures or when they are looking at sentences. An interesting difference between many applications of machine learning, and this study is that most machine learning algorithms have the goal of achieving the maximum possible discrimination between conditions or classes. However, for fMRI studies it is important that the pattern of activity that is identified by the algorithm is a pattern associated with the cognitive state. When using non-linear classifiers, it is possible to decode features predictive of the cognitive state, where the brain does not actually respond to these features. For example, when decoding the primary visual cortex with high enough spatial resolution, it would be possible to combine simple features with a non-linear neural network to decode the presence or absence of, for example, faces. However, this information is not actually processed in that part of the brain \cite{to:2012rev}. To avoid this problem it is recommended to only use linear classifiers when decoding cognitive states from fMRI data. The classifiers that we compare in this paper are Gaussian Naive Bayesian (GNB) classifiers, Linear Support Vector Machine (LSVM), and a Multivariate Gaussian (MVG) classifier.

The goal of this study is to compare the classification performance of the cognitive states associated with looking at a picture or looking at a sentence between subject. This objective introduces an important issue, namely, how the data from the fMRI scans can be compared between subjects. There are two mains ways this could be achieved. First, groups of voxels that are anatomically defined as belonging to a certain brain region, a region of interest (ROI), can be averaged into a single \emph{supervoxel}, which can be compared between subjects. Second, an anatomically defined talairach coordinate system can be used that maps the voxels of each subject onto an anatomically defined coordinate. The ROI averaging it is a rather crude method, and a lot of information of the individual voxels is lost. However, because the ROI's are relatively easy to identify by an expert they can be compared easily between subjects. By using the talairach coordinate system the data of each individual voxel can be used. However, because the error of doing this is usually in the order of a centimetre, while the size of the voxels are in the order of millimetres, the coordinates are difficult to generalise. Because we aim to do between-subjects comparisons the generalisation performance is important, thus we will use averaged ROI's in order to compare the classifiers \cite{mi:2004coord}.\\

@@

ROI mapping. ROI mapping is a form of feature selection, wherein only a subset of the d-dimensional input space is used. As described in the introduction, ROIs are anatomically defined brain areas classified as relevant by an expert in the field. To reduce the number of features, average voxel values within this regions are taken as input features. As other higher-order features such as the standard deviation, kurtosis and skewness within ROIs did not increase predictability, these were not included in the subsequent analysis.

Principal Component Analysis (PCA). As feature extraction method, we applied PCA. In between subject classification, the input vectors of the PCA should ideally be the standardized Talairach coordinates. As this coordinate system appeared to be inaccurate for this dataset, we were forced to use the voxel indices. Because brain sizes and shapes differ per subject, it is uncertain whether the voxel indices map well onto each other. As a result, the accuracy of the PCA is reduced.

>>>>

%% Abstract section.
\abstract{
Functional magnetic resonance imaging (fMRI) is a brain mapping technique that allows researchers to infer activity in a subject's brain. This technique is typically used to determine which areas of the brain are involved in performing a certain task. Using this knowledge, we may also be able to determine what task was performed by looking at fMRI data. In this paper, we will compare machine learning algorithms in between-subjects classification performance of cognitive states using fMRI data. We found that averaging across ROI gave the best classification performance. All classifiers that we used performed above-chance, and we found that a Multivariate Gaussian classifier outperformed a Naive Bayesian classifier and a Support Vector Machine. Overall we can say that machine learning algorithms provide a feasible method for decoding cognitive states from fMRI data.
} % end of abstract

@@

In this study we use the same StarPlus data as in the previously mentioned experiment, and use machine learning algorithms to decode the cognitive states of subjects when they are looking at pictures or when they are looking at sentences. An interesting difference between many applications of machine learning, and this study is that most machine learning algorithms have the goal of achieving the maximum possible discrimination between conditions or classes. However, for fMRI studies it is important that the pattern of activity that is identified by the algorithm is a pattern associated with the cognitive state. When using non-linear classifiers, it is possible to decode features predictive of the cognitive state, where the brain does not actually respond to these features. For example, when decoding the primary visual cortex with high enough spatial resolution, it would be possible to combine simple features with a non-linear neural network to decode the presence or absence of, for example, faces. However, this information is not actually processed in that part of the brain \cite{to:2012rev}. To avoid this problem it is recommended to only use linear classifiers when decoding cognitive states from fMRI data. The classifiers that we compare in this paper are Gaussian Naive Bayesian (GNB) classifiers, Linear Support Vector Machine (LSVM), and a Multivariate Gaussian (MVG) classifier.\\
\indent The goal of this study is to compare the classification performance of the cognitive states associated with looking at a picture or looking at a sentence between subject. This objective introduces an important issue, namely, how the data from the fMRI scans can be compared between subjects. There are two mains ways this could be achieved. First, groups of voxels that are anatomically defined as belonging to a certain brain region, a region of interest (ROI), can be averaged into a single \emph{supervoxel}, which can be compared between subjects. Second, an anatomically defined talairach coordinate system can be used that maps the voxels of each subject onto an anatomically defined coordinate. The ROI averaging it is a rather crude method, and a lot of information of the individual voxels is lost. However, because the ROI's are relatively easy to identify by an expert they can be compared easily between subjects. By using the talairach coordinate system the data of each individual voxel can be used. However, because the error of doing this is usually in the order of a centimetre, while the size of the voxels are in the order of millimetres, the coordinates are difficult to generalise. Because we aim to do between-subjects comparisons the generalisation performance is important, thus we will use averaged ROI's in order to compare the classifiers \cite{mi:2004coord}.\\

@@

ROI mapping. As described in the introduction, ROIs are anatomically defined brain areas classified as relevant by an expert in the field. To reduce the number of features, average voxel values within this regions are taken as input features. As other higher-order features such as the standard deviation, kurtosis and skewness within ROIs did not increase predictability, these were not utilized.

Principal Component Analysis (PCA).
